---
title: "DSC365: Introduction to Data Science"
author: "Principal Component Analysis"
date: "April 2, 2026"
output: 
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r echo=FALSE, message=FALSE, warning = FALSE}
library(tidyverse)
library(knitr)
library(RColorBrewer)
library(mosaic)
library(infer)
library(Metrics)


hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = xfun::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})

```

### Announcements

**Mini Project 3**

- Due April 9th at 3:30pm in Blueline (peer review during class)

**Lab 6** 

- Due Tuesday April 7th at 11:59 pm in Blueline

<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>

[Modern Data Science with R, Chapter 12](https://mdsr-book.github.io/mdsr3e/12-learningII.html)

[Interactive Walk Through](https://bryanhanson.github.io/LearnPCA/articles/Vig_05_Visualizing_PCA_3D.html)

---
### Changes to Rest of Semester Schedule

(Also posted in Blueline)


| Tuesday  | Topic               | Thursday | Topic        | Assignment Due                  |
| :------- | :-----------------: | :------- | :----------: | :-----------------------------: |
| April 7  | Text Data           | April 9  | MP3/Lab 7    | April 7 - Lab 6, April 9 - MP3  |
| April 14 | Spatial Data        | April 16 | No Class     | April 16 - Lab 7                |
| April 21 | Lab 8               | April 23 | Web-Scraping |                                 |
| April 28 | Quiz 4, Lab 9       | April 30 | Lab 9        | April 28 - Lab 8, May 1 - Lab 9 |  
| May 5    | MP4 Presentations   |          |              | MP4 Report/Presentation - May 5 |


---
### Principal Component Analysis (PCA): Why?

Sometimes a data set has a lot of variables. Some of these variables carry *lots* of information. Others carry *little to no* information. PCA can help reduce the dimensions of data by summarizing most of the variation into fewer dimensions called components. 

<br>

---
### Principal Component Analysis (PCA): Why?

**PCA**:

---
### PCA: Overview of Steps

1).

<br>
<br>
<br>
<br>
<br>


2). 

---
### PCA: Overview of Steps

3). Find Principal Components
  + First Principal Component of a set of features $X_1, \dots X_p$ is a normalized linear combination of the features: $Z_1 = \phi_{11}X_1 + \phi_{21}X_2 + \cdots + \phi_{p1}X_p$ that has the largest variance.
    - Normalized: $\sum^p_{j=1} \phi^2_{j1} = 1$
    - To find the loadings, solve the optimization problem: $\text{maximize}_{\phi_{11}, \dots \phi_{p1}}(\frac{1}{n}\sum^{n}_{i=1}(\sum^{p}_{j=1}\phi_{j1}x_{ij})^2)$

<br>
<br>
<br>
<br>
<br>

  + Second Principal Component: maximal variance out of all linear combinations that are uncorrelated (orthogonal) with $Z_1$
      
---
### PCA: Picture!


---
### PCA: Overview of Steps

4). Select the number of components that capture most of the variation

  
  
  
---
### Example: Data

The Happy Planet Index is a measure of a country’s ecological and societal “well-being”.

```{r}
library(readxl)
happyplanet2016 <- read_excel("happyplanet2016.xlsx")
glimpse(happyplanet2016)
```

Can we distill this down to a few important variables or ideas?

---
### Example: Fitting a PCA

The Happy Planet Index is calculated using a weighted average of these variables. It wouldn’t make sense to try to find linear combinations with a variable that’s already a linear combination, so we remove it. 

How are the other variables related?


```{r}
happyplanet2 <- na.omit(happyplanet2016[,-8])

#scale and find covariance matrix as part of function
happy_pca <- prcomp(x=happyplanet2[,3:10], 
                    center=TRUE, scale=TRUE) 

summary(happy_pca) #proportion of variance explained
```

---
### Fitting a PCA: Loadings

```{r}
happy_pca$rotation #loadings
```

---
### Fitting a PCA: Number of Components

How many components should we choose?

```{r, message=FALSE, fig.align='center', fig.width= 12, fig.height=5}
#install.packages("factoextra")
library(factoextra)
fviz_eig(happy_pca, main = "Scree plot: Happy Planet Index")
```


---
### Fitting a PCA: Biplot

Can also create a bi-plot - provides visualization of PCA outputs

+ Perpendicular vectors:

<br>
+ Two variables in the same direction:

<br>
+ Two variables in the opposite direction:

---
### Fitting a PCA: Biplot

```{r, eval = FALSE}
fviz_pca_var(happy_pca, col.var = "contrib")
```

```{r, echo = FALSE, fig.align='center', fig.height=10, fig.width=16}
fviz_pca_var(happy_pca, col.var = "contrib", labelsize = 8) +
  theme(axis.title = element_text(size = 25, face = "bold"))
```




---
### Fitting a PCA:  `ggbiplot`

If a package you might want to use  is on GitHub and you know the user’s Git “repository”: install the package using the `install_github` function in the `devtools` library.

```{r, message=FALSE}
library(devtools)
#install_github("vqv/ggbiplot")
library(ggbiplot)
```


```{r, eval = FALSE, fig.align='center', fig.height=8, fig.width=16}
ggbiplot(happy_pca, groups=happyplanet2$Region, ellipse=TRUE)
```

```{r, fig.align='center', echo=FALSE, fig.height=3}
ggbiplot(happy_pca, groups=happyplanet2$Region, ellipse=TRUE, labels.size = 15) +
  theme(axis.title = element_text(size = 15, face = "bold"))
```

---
### Fitting a PCA:  `ggbiplot`

Can also visualize other PCs besides the first two.

```{r, fig.align='center', fig.height=8, fig.width=19, eval = FALSE}
ggbiplot(happy_pca, groups=happyplanet2$Region, ellipse=TRUE, 
         choices=c(3,4))
```

```{r, fig.align='center', fig.height=8, fig.width=19, echo = FALSE}
ggbiplot(happy_pca, groups=happyplanet2$Region, ellipse=TRUE, 
         choices=c(3,4), labels.size = 10) +
  theme(axis.title = element_text(size = 25, face = "bold"))
```


---

### In Summary

Advantages:

<br>
<br>
<br>
<br>
<br>
<br>
<br>

Disadvantages:


---

#### Try Yourself

Try the PCA with the gapminder data. Use all four variables (`infant_mortality`, `life_expectancy`, `fertility` and `popualtion`) try to make a data visualization to show the relations between the variables.

```{r}

```


